---
title: "Differential prevalence analysis with DiPPER"
author:
  - name: Juho Pelto
    affiliation: University of Turku
    email: jepelt@utu.fi
  - name: Leo Lahti
    affiliation: University of Turku
    email: leo.lahti@iki.fi
package: DiPPER
output:
  BiocStyle::html_document:
    toc_float: true
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Differential Prevalence Analysis with DiPPER}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

# Checks whether CmdStan is available and the model-fitting chunks can be
# evaluated only when CmdStan is present. Otherwise the pre-computed fits in
# inst/extdata are used.
cmdstan_ok <- !identical(Sys.getenv("DIPPER_NO_CMDSTAN"), "true") &&
    requireNamespace("instantiate", quietly = TRUE) &&
    instantiate::stan_cmdstan_exists()

cache_file <- function(x) {
    system.file("extdata", x, package = "DiPPER")
}

have_cache <- nzchar(cache_file("fit_hintikka.rds")) &&
    nzchar(cache_file("fit_vatanen.rds"))

# Run the DiPPER models?
run_fit <- cmdstan_ok

# Can results be shown?
show_res <- cmdstan_ok || have_cache
```

# Introduction

Recent research suggests that presence/absence information of taxonomic
features (e.g., species or genera) alone, rather than complete abundance
profiles, can be highly informative when analyzing associations between
microbiomes and external variables such as health outcomes
[@Giliberti2022HostTaxab; @Karwowska2025EffectsData]. In particular,
*differential prevalence analysis (DPA)*, which in its simplest form compares
feature prevalences between two groups (e.g., healthy vs. non-healthy
subjects), represents a compelling complementary or even alternative approach
to traditional differential abundance analysis
[@Pelto2025ElementaryAnalysis; @Nickols2026MaAsLinDiscovery].

DiPPER (**Di**fferential **P**revalence via **P**robabilistic **E**stimation
in **R**) is a Bayesian hierarchical model-based approach designed for
differential prevalence analysis. In addition to the main variable of interest
(e.g., host disease status), it can incorporate other *covariates* (e.g., age,
sex, BMI). It can also be used to analyze *longitudinal* (or repeated measures)
microbiome data.

## Advantages of DiPPER

While differential prevalence analysis can be performed using approaches based
on standard (frequentist) logistic regression, such approaches suffer certain
limitations that DiPPER aims to solve:

* **Robustness in boundary cases:** Boundary cases (or quasi-complete
  separation), where a taxon is completely absent or fully present in one
  study group, are common in microbiome data. Standard approaches may have
  difficulties in these cases. For instance, the default logistic regression
  implementation in R (based on the Wald test) yields non-sensible *p*-values
  and infinite parameter estimates and confidence intervals in boundary cases.
  DiPPER, however, remains robust in these cases, providing finite estimates
  and uncertainty intervals.

* **Applicability to longitudinal data:** The aforementioned boundary cases
  can be even more problematic in longitudinal data modeling. While there are
  robust versions of frequentist logistic regression (e.g., Firth's penalized
  regression [@Firth1993BiasEstimates]), they can be difficult to generalize
  to longitudinal settings. DiPPER, however, provides a single, unified
  robust framework for both cross-sectional and longitudinal analyses.

* **Inherent multiplicity adjustment:** Standard approaches typically calculate
  an unadjusted *p*-value for each feature and subsequently apply a post-hoc
  adjustment (e.g., the Benjamini-Hochberg procedure) to control for multiple
  testing (while leaving the corresponding regression coefficients and
  confidence intervals unadjusted). In contrast, multiplicity adjustment in
  DiPPER occurs naturally via hierarchical shrinkage. The differential
  prevalence parameters are assumed to originate from a common (Asymmetric)
  Laplace prior distribution. This shrinks the estimates toward zero and toward
  each other, providing effectively multiplicity-adjusted estimates and
  intervals that can directly be used to assign "statistical significance".

## Benchmarking results against alternative methods

We have benchmarked DiPPER against alternative methods using real data from
57 human gut microbiome studies. The performance was compared against three
versions of frequentist logistic regression, DPA implementations in the
MaAsLin3 [@Nickols2026MaAsLinDiscovery] and LDM [@Hu2021AMicrobiome] packages,
and two differential abundance methods (MaAsLin2
[@Mallick2021MultivariableStudies] and LinDA [@Zhou2022LinDA:Data]).

DiPPER demonstrated high sensitivity in detecting potentially differentially
prevalent features while maintaining a well-calibrated family-wise error rate
under the global null hypothesis. Moreover, it outperformed all
alternatives in the replication of findings across independent studies.

## More information

A detailed description of DiPPER, along with comprehensive benchmarking
results, can be found in our pre-print [@Pelto2026DiPPER:Studies].

# Installation

Installing DiPPER is generally straightforward, but it requires dependencies
from both Bioconductor (for handling microbiome data structures) and Stan (for
Bayesian inference).

## Stan backend (cmdstanr)

Unless you already have `cmdstanr` installed, you will need to set it up first.
While configuring the Stan backend can occasionally require some
troubleshooting depending on your operating system, following these steps
should generally ensure a smooth setup.

First, install the `cmdstanr` R package from the Stan R-universe repository.

```{r install_stan, eval=FALSE}
# Install cmdstanr package
install.packages("cmdstanr",
                 repos = c("https://stan-dev.r-universe.dev",
                           getOption("repos")))

# Check C++ toolchain and install the backend
cmdstanr::check_cmdstan_toolchain(fix = TRUE)
cmdstanr::install_cmdstan()
```

## Bioconductor dependencies and DiPPER

Once CmdStan is ready, you can install the development version of DiPPER.
Using `BiocManager` for the installation automatically installs all required
Bioconductor dependencies.

*Note: DiPPER compiles its Stan models during installation, which may take a
few minutes and produce verbose output. This is expected, and it is done only
once: afterwards the models are ready to use.*

```{r install_bioc, eval=FALSE}
# Install BiocManager if it is not already installed
if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
}

# Install DiPPER from GitHub
BiocManager::install("jepelt/DiPPER")
```


# Quick start

Below is a minimal workflow using the built-in `tse_hintikka` dataset, which
compares the gut microbiomes of rats (N = 20 + 20) on a high-fat versus low-fat
diet (`Fat`) while controlling for xylo-oligosaccharide (`XOS`)
supplementation. The dataset originates from
[Hintikka et al. (2021)](https://www.mdpi.com/1660-4601/18/8/4049).

We run DiPPER with its default settings. We only need to provide the dataset,
the model formula, and the name of the assay containing the count data
(`assay.type = "counts"`). Note that the *variable of interest* (here `Fat`)
must be set as the *first* variable in the model formula. Note also that DiPPER
by default automatically computes and controls for the library size (sequencing
depth) in the model.

```{r quick_start, eval=run_fit}
library(DiPPER)

# Load dataset (TreeSummarizedExperiment object)
data("tse_hintikka")

# Run DiPPER with the default settings
fit <- dipper(
    tse = tse_hintikka,
    formula = ~ Fat + XOS,
    assay.type = "counts"
)
```

```{r quick_start_cached, eval=!run_fit && have_cache, echo=FALSE}
# CmdStan is not available on this machine: use the pre-computed fit instead
library(DiPPER)
fit <- readRDS(cache_file("fit_hintikka.rds"))
```

Next, we extract the summarized results as a `data.frame` and print the first
seven rows (the most "significant" genera). Note that the data frame contains
the results only for the variable of interest, which in this case is the `Fat`
variable indicating the diet, low-fat ('Low', the reference level) or high-fat
('High').

```{r quick_results, eval=show_res}
# Extract the results
res <- summary(fit)

# Print the first seven rows
res[1:7, ]
```

The summarized results `data.frame` contains the following columns:

* **`taxon`**: Taxon name (here: genus).
* **`log_or`**: Differential prevalence estimate *for the variable of interest*
  (here: `Fat`), expressed as a log(odds ratio).
* **`lwr`**, **`upr`**: Lower and upper limits of the 95% credible interval for
  the differential prevalence.
* **`significant`**: A logical value indicating whether the 95% credible
  interval excludes zero. This roughly corresponds to traditional "statistical
  significance" based on multiplicity-adjusted *p*-values.
* **`pseudo_q`**: Indicates the highest "significance level" at which the
  result would be considered "significant" (i.e., the credible interval
  excluding zero). It thus bears some resemblance to a multiplicity-adjusted
  two-sided *p*-value (*q*-value), but it does not have any strict statistical
  interpretation.

Lastly, create a forest plot (with 95% credible intervals) showing the
significant genera (determined by the 95% credible intervals). Here, the odds
ratio (OR) values > 1 indicate higher prevalence in the high-fat diet group.

```{r quick_plot, eval=show_res, out.width="100%"}
plot(fit)
```

# More complex example

This example demonstrates a more complex case, where we have longitudinal data.

We use a built-in subset of
the [Vatanen et al. (2016)](https://pmc.ncbi.nlm.nih.gov/articles/PMC4950857/)
dataset. The dataset includes gut microbiome samples from 79 infants at one
or two time points (1.5 and/or 7 months of age). We want to study how
microbe prevalences depend on age point (1.5 months vs 7 months)
while controlling for gender and antibiotics use.

```{r load_vatanen}
library(DiPPER)

# Load dataset (TreeSummarizedExperiment object, agglomerated to genus level)
data("VatanenT_2016_subset")
```

The metadata includes the variables `subject_id` (indicating the infant),
`age_point` (1.5 months or 7 months), `antibiotics` (use of antibiotics:
yes/no), and `gender` (male/female). Note that there are 1-2 measurements per
infant in the data.

```{r vatanen_metadata}
VatanenT_2016_subset |>
    SummarizedExperiment::colData() |>
    head()
```

Note that the abundance data are now proportions (relative abundances):

```{r vatanen_abundance}
vatanen_assay <- SummarizedExperiment::assay(VatanenT_2016_subset,
                                             "relative_abundance")
vatanen_assay[c("Bifidobacterium", "Escherichia", "Veillonella"), 1:5]
```

As we want to study the effect of age point (1.5 months vs 7 months) on the
microbe prevalences, we run DiPPER by setting `age_point` as the first variable
in the formula, making it the variable of interest. As we control for
`antibiotics` use and `gender`, we put them as the second and third variables.
Lastly, as some infants have two measurements, we must
account for the longitudinality by specifying `subject_id` (indicating infant)
as a random intercept term: `(1 | subject_id)`. Our formula is thus
`~ age_point + antibiotics + gender + (1 | subject_id)`.

Furthermore, because the abundance data are proportions (and not counts), we
must set `data.type = "relabundance"`. Moreover, because our data does not
contain sequencing (read) depth information, we cannot control for it in the
model, and must thus set `read.depth = FALSE`.

Lastly, we set `niter = 1000`. This argument tells the number of MCMC samples
used to approximate the posterior distribution of the model parameters. A
higher value gives more accurate estimates, but also increases the computation
time. The default `niter = 2000` should be enough in almost any case, but
`niter = 1000` is often sufficient. A potentially too low
number of MCMC iterations would be indicated by the model diagnostics.

```{r assay_vatanen, eval=run_fit}
fit_long <- dipper(
    tse = VatanenT_2016_subset,
    formula = ~ age_point + antibiotics + gender + (1 | subject_id),
    assay.type = "relative_abundance", # Name of the assay
    data.type = "relabundance",
    read.depth = FALSE,
    niter = 1000
)
```

```{r assay_vatanen_cached, eval=!run_fit && have_cache, echo=FALSE}
# CmdStan is not available on this machine: use the pre-computed fit instead
fit_long <- readRDS(cache_file("fit_vatanen.rds"))
```

```{r diagnostics_note, eval=run_fit, echo=FALSE, results="asis"}
# Shown only when the model was actually fitted above, as the diagnostic
# message is not printed when a pre-computed fit is loaded instead.
cat(
    "Note that the model diagnostics (\"All MCMC diagnostics are within",
    "acceptable limits.\") indicate that the MCMC sampling has converged",
    "well, and that `niter = 1000` was thus enough.\n"
)
```

Finally, we extract the results (for `age_point`) and create a forest plot
showing only the significant taxa. As expected, the prevalence of several taxa
increases with infant age, although the prevalence of *Staphylococcus*
decreases.

```{r plot_vatanen, eval=show_res, out.width="100%"}
# Extract the results for age_point
res2 <- summary(fit_long)

# Create a forest plot for significant taxa
plot(fit_long)
```

# Session Information

```{r session_info}
sessionInfo()
```
